A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-09-08 bioc-workflow.Rmd
added tlres from pRolocdata

The first step, as with any machine learning algorithm, is to optimise
any free paramaters of the classifier. For the *k*-NN TL classifier
there are two sets of parameters that need optimising: the first set
are the $k$'s for the primary and auxiliary data sources required for
the nearest neighbour calculations for each data source.

```{r knnoptfortl, eval=FALSE}
## get best k's
kpopt <- knnOptimisation(hl, fcol = "tlmarkers")
kaopt <- knnOptimisation(gocc, fcol = "tlmarkers")
kp <- getParams(kpopt)
ka <- getParams(kaopt)
```

The second
set of parameters (which is noted by a vector of $\theta$ weights)
that require optimising are the class weights, one per subcellular
niche, that control the proportion of primary and auxiliary data to
use for learning. A weight can take any real value number between 0
and 1. A weight of $\theta = 1$ indicates that all weight is given to
the primary data (and this implicitly implies that a weight of $1 -
\theta$ is given to the auxiliary data, similarly a weight of $\theta
= 0$ implies that all weight is given to the auxiliary data (so 0 is
given to the primary source). If we conduct a parameter seach and test
weights $\theta = {0, 1/3, 2/3, 1}$ for each class, and if we have,
for example 10 subcellular niches, this will result in `4^10`
different combinations of parameters to test. The parameter
optimisation is therefore time consuming and as such we recommend you
make use of a computing cluster (code and submissing scripts are also
available in the supporting information). The markers in the `hl`
dataset contain 14 subcellular classes. If we examine these markers
and classes on the PCA plot above we can see that in particular the
two ribosomes and two nuclear compartments are highly separated along
the first two components, this is also evident from the profiles plot
which gives us a good indication that these subcellular niches are
well-resolved in the hyperLOPIT dataset. Transfer learning is
particularly useful for classes that are not as well separated, we
find that subcellular niches that are well-separated under hyperLOPIT
and LOPIT obtain a class score of 1 (i.e. use only primary data from
transfer learning [@Breckels:2016]). Therefore, for the optimisation
stage of the analyses we can already infer a subcellular class weight
of 1 for these niches and only optimise over the remaining organelles.
This can signifciantly cut down optimisation time as by removing these
4 classes from the optimisation (and not the classification) we only
have `4^10` class weight combinations to consider instead of `4^14`
combinations. In the example below we remove these 4 classes from the
marker set, re-run the `knnOptimisation` for each data source and then
run the `knntlOptimisation` with the 10 remaining classes. (Note: this
is not run live as this the `hl` dataset with 10 classes, 707 markers
and `4^10` combinations of parameters takes ~76 hours to run on the
University of Cambridge HPC using 256 workers).

In the example code below we first remove the 2 ribosomes and nuclear
classes and then run the `knnOptimisation` followed by the
`knntlOptimisation` (specifying snow with a MPI
backend with 255 workers for parallelisation on the Cambridge HPC).

```{r loadtlres, include = FALSE}
load("Data/tl_runs/tlopt.rda")
data("hyperLOPIT2015goCC")
gocc <-hyperLOPIT2015goCC
ka <- kp <- 3
```

```{r setuptlopt, eval=TRUE}
## create new markers column for tl markers
fData(hl)$tlmarkers <- fData(hl)$markers
fData(gocc)$tlmarkers <- fData(gocc)$markers

## Remove 4 classes
torm  <- c("40S Ribosome", "60S Ribosome",
           "Nucleus - Chromatin", 
           "Nucleus - Non-chromatin")
for (i in seq(torm)) {
  hl <- fDataToUnknown(hl, from = torm[i], fcol = "tlmarkers")
  gocc <- fDataToUnknown(hl, from = torm[i], fcol = "tlmarkers")
}
getMarkerClasses(hl, fcol = "tlmarkers")
getMarkerClasses(hl, fcol = "tlmarkers")
```
```{r runtlopts, eval=FALSE}
## get best k's
kp <- knnOptimisation(hl, fcol = "tlmarkers")
ka <- knnOptimisation(gocc, fcol = "tlmarkers")

## Set appropriate backend and workers for the tl
par <- SnowParam(255L, type = "MPI")

@@

```{r plottl}
plot(tlopt)
```

>>>>

The first step, as with any machine learning algorithm, is to optimise
any free paramaters of the classifier. For the *k*-NN TL classifier
there are two sets of parameters that need optimising: the first set
are the $k$'s for the primary and auxiliary data sources required for
the nearest neighbour calculations for each data source. The second
set of parameters (which is noted by a vector of $\theta$ weights)
that require optimising are the class weights, one per subcellular
niche, that control the proportion of primary and auxiliary data to
use for learning. A weight can take any real value number between 0
and 1. A weight of $\theta = 1$ indicates that all weight is given to
the primary data (and this implicitly implies that a weight of $1 -
\theta$ is given to the auxiliary data, similarly a weight of $\theta
= 0$ implies that all weight is given to the auxiliary data (so 0 is
given to the primary source). If we conduct a parameter seach and test
weights $\theta = {0, 1/3, 2/3, 1}$ for each class, and if we have,
for example 10 subcellular niches, this will result in `4^10`
different combinations of parameters to test. The parameter
optimisation is therefore time consuming and as such we recommend you
make use of a computing cluster (code and submissing scripts are also
available in the supporting information). The markers in the `hl`
dataset contain 14 subcellular classes. If we examine these markers
and classes on the PCA plot above we can see that in particular the
two ribosomes and two nuclear compartments are highly separated along
the first two components, this is also evident from the profiles plot
which gives us a good indication that these subcellular niches are
well-resolved in the hyperLOPIT dataset. Transfer learning is
particularly useful for classes that are not as well separated, we
find that subcellular niches that are well-separated under hyperLOPIT
and LOPIT obtain a class score of 1 (i.e. use only primary data from
transfer learning [@Breckels:2016]). Therefore, for the optimisation
stage of the analyses we can already infer a subcellular class weight
of 1 for these niches and only optimise over the remaining organelles.
This can signifciantly cut down optimisation time as by removing these
4 classes from the optimisation (and not the classification) we only
have `4^10` class weight combinations to consider instead of `4^14`
combinations.

In the example below we remove these 4 classes from the
marker set, re-run the `knnOptimisation` for each data source and then
run the `knntlOptimisation` with the 10 remaining classes. (Note: this
is not run live as this the `hl` dataset with 10 classes, 707 markers
and `4^10` combinations of parameters takes ~76 hours to run on the
University of Cambridge HPC using 256 workers).

To remove the 4 classes and create a new column of markers in the feature data called `tlmarkers` to use for the analysis:
```{r setuptlopt, eval=TRUE}
## create new markers column for tl markers
fData(hl)$tlmarkers <- fData(hl)$markers
fData(gocc)$tlmarkers <- fData(gocc)$markers

## Remove 4 classes
torm  <- c("40S Ribosome", "60S Ribosome",
           "Nucleus - Chromatin", 
           "Nucleus - Non-chromatin")
for (i in seq(torm)) {
  hl <- fDataToUnknown(hl, from = torm[i], fcol = "tlmarkers")
  gocc <- fDataToUnknown(hl, from = torm[i], fcol = "tlmarkers")
}
getMarkerClasses(hl, fcol = "tlmarkers")
getMarkerClasses(hl, fcol = "tlmarkers")
```

Optimisation stage 1: run `knnOptimisation` to get the best $k$'s for each data source.
```{r knnoptfortl, eval=FALSE}
## get best k's
kpopt <- knnOptimisation(hl, fcol = "tlmarkers")
kaopt <- knnOptimisation(gocc, fcol = "tlmarkers")
kp <- getParams(kpopt)
ka <- getParams(kaopt)
```

Optimisation stage 2: run `knntlOptimisation` to get the best transfer learning weights for each sub-cellular class.
```{r runtlopts, eval=FALSE}
## Set appropriate parallelisation backend and  
## number of workers for the tl
par <- SnowParam(255L, type = "MPI")

@@

```{r loadtlres, echo=FALSE}
tlfile <- dir(extdatadir, full.names = TRUE,
              pattern = "bpw-tlopt.rds")
tlopt <- readRDS(tlfile)

```
```{r plottl}
plot(tlopt)
```
